{\bf \huge Wikiplia}:\\ The Free Programming Language\\ That Anyone Can Edit

### 2007-03-19 trunk/wikiplia/paper/wikiplia.tex
almost there...

\newcommand\ftinker{\ensuremath{\copyright}}
\newcommand\fexpression{\ensuremath{\hbar}}
\newcommand\fbeer{\ensuremath{\triangle}}
\newcommand\fusa{\ensuremath{\sqrt[x^{1.2}]{\mbox{\t{oo}} + \frac{x}{z^2}}}}
\newcommand\ffree{\ensuremath{\P}}

@@

\vspace{1em}
{\noindent \small {\bf Keywords}:
  Freedom, programming language, software license, wiki, XML, 
  hyper-driven devices
}

@@

@@

\caption{Evaluation of XML, part 2. The {\sf rate} judgment assesses a
sequence of XML expressions. The rules for rating the primitives {\sf
string}, {\sf sub}, {\sf substr}, {\sf +}, {\sf -}, {\sf int}, {\sf
eq} and {\sf if} are omitted for space.
%
} \label{fig:xmlrate}
\end{figure*}

@@

@@

>>>>

\newcommand\ftinker{\ensuremath{\copyright}}
\newcommand\fexpression{\ensuremath{\hbar}}
\newcommand\fbeer{\ensuremath{\triangle}}
\newcommand\fusa{\ensuremath{\sqrt[x^{1.2}]{\mbox{\t{oo}} + \frac{x}{z^2}}}}
\newcommand\ffree{\ensuremath{\P}}
\newcommand\ftobe{{\tiny \ensuremath{\stackrel{\bigcirc}{\amalg}}}}

@@

\vspace{1em}
{\noindent \small {\bf Keywords}:
  Freedom, programming language, software license, wiki, XML, 
  weatherproof footwear and their fastening mechanisms,
  hyper-driven devices,
}

@@

\paragraph{Freedom \ftobe: Free to be You and Me.}
The development of a programming language should not be confined to
the bearded academic elite, gazing down upon the programmer fiefs from
their stratospheric ivory towers. Wikiplia is the free programming
language that {\em anyone} can edit: from bearded academic
elite\wcite{Tenure} to congressional staffers\wcite{Congressional
staffer edits to Wikipedia} to nameless Slashdot\wcite{Slashdot}
trolls. Wikiplia's WWW-based approach means that an Internet
connection and compatible WWW hypertext browser is all that's needed
to begin on the life-changing journey of programming language design.

@@

\caption{Evaluation of XML, part 2. The {\sf rate} judgment assesses a
sequence of XML expressions. The rules for rating the primitives {\sf
parse}, {\sf string}, {\sf sub}, {\sf substr}, {\sf +}, {\sf -}, {\sf
int}, {\sf eq} and {\sf if} are omitted for space.
%
} \label{fig:xmlrate}
\end{figure*}

@@

The system is implemented as a Standard~ML\wcite{Standard ML} program
of approximately 1,000 lines.\z\footnote{This count does not include
general purpose libraries, such as a networking library.} This program
is designed to be minimal: it consists of a web server, a revision
control system, and facilities for evaluating XESP documents. It also
contains a very minimal bootstrapping ``compiler'' for XESP documents,
with its boots manually strapped. From this tiny core we then develop
the remainder of Wikiplia using Wikiplia itself.

Some may balk at the choice of Standard~ML, as the language is
miserably non-free: First, while many of the major implementations are
GNU or BSD-licensed,\z\wcite{BSD licenses} all are implemented in
Standard~ML itself, yielding the bootstrapping problem described
earlier. Second, the mathematical Definition of Standard ML book is
only available as a copyrighted publication of MIT
Press,\z\wcite{MIT Press} not even {\em ostensibly} in a free
manner. However, the performance considerations of the server and
evaluator---and the lack of suitable free alternatives---force us to
settle for such subjugation.

\subsubsection{Web Server}

The web server's job is simple. It runs in a loop, accepting a single
connection, setting up an initial environment $\G$ for evaluation
(which associates the symbols {\sf request.url}, {\sf request.ip} and
{\sf request.time} with the appropriate values), and then evaluating
the document contained at the head of the key ``main'' in the
repository. The result of that evaluation is sent back to the web
browser as a string.\z\footnote{The HTTP result code may be modified
if, for example, the resulting document is a HTTP redirect to another
URL.} The server knows nothing else about how Wikiplia works.

The web server is single-threaded (so each request must finish
before the next is handled) because we desire global atomicity
(Section~\ref{sec:bootstrap}).

\subsubsection{Revision Control}

The revision control system stores XESP documents and their history.
This is mostly a straightforward implementation of the imperative {\tt
cvs} judgments given in Section~\ref{sec:calculus}. Unlike typical
revision control systems, we need to support very large numbers of
revisions with small edits (since every change is saved), so the
implementation is engineered to make storing small revisions very
cheap. Particularly, revisions are aggressively compressed by only
storing the newest revision directly, and then a series of difference
``plans'' that describe how to get the next older revision from the
current one. We compute optimal plans using an efficient minimal edit
distance\wcite{Levenshtein distance} calculation at the
token\wcite{Token (parser)} level. As of revision~\currentrevision,
the database is only 236 kilobytes.\z\wcite{Kilobyte}

\subsubsection{Document Evaluation}

The Wikiplia implementation also has facilities for parsing and
evaluating XESP documents. These are a direct implementation of the
evaluation rules in Figures~\ref{fig:xmleval}~and~\ref{fig:xmlrate}.

\subsubsection{Bootstrapping}

The key ``main'' of Wikiplia is responsible for decomposing the URL
and acting upon it however is appropriate. The goal of the
bootstrapping process is to make Wikiplia
self-sufficient\wcite{Self-sufficiency} in the sense that the language
can be edited from the web site implemented by ``main.'' To do so, the
web site needs to be able to present the user with an edit box
containing the current source of the ``main'' key and the ability to
save his code into the database, overwriting the ``main'' key, in
order to add functionality.

The initial implementation of ``main'' is provides for the simple
ability to edit, save, and view the current version of keys in the
repository. These three actions are encoded as the URLs {\tt
/edit/key}, {\tt /save/key}, and {\tt /view/\_/key}. The view action
is straightforward. The edit action displays an HTML
textarea\wcite{Text box} containing the current value of the key and a
button that submits the changes to the {\tt save} url. The save action
is the most complex. First, the submitted document is saved as the new
version of the key. Then, if the key is of the form {\tt
base.extension}, the database is checked to see if there is a key
called {\tt extension:compile}. If so, the XESP document that is there
is applied to the input document (a string) to produce a document that
is saved at the key {\tt base}. This allows us to develop languages
that are automatically compiled when saved.

\begin{figure}
\begin{center}
\verb+(lambda 's '(parse (xcase s 'no '(h _ h))))+
\end{center}
\caption{The initial bootstrapping compiler.} \label{fig:b:compile.b}
\end{figure}

This initial functionality is implemented directly in the XESP
language, whose extension is {\tt b}; the bootstrapping ``compiler''
is just the built-in parser (Figure~\ref{fig:b:compile.b}).

Arranging that the repository contain the correct keys to make this
work is subtle. A small initialization phase sets up:
\begin{itemize}
\item {\tt b:compile.b}$\quad$ 
      The source code (a string) in Figure~\ref{fig:b:compile.b}
\item {\tt b:compile}$\quad$ 
      The parsed document corresponding to the above, such that {\tt b:compile} applied to {\tt b:compile.b} yields {\tt b:compile}
\item {\tt main.b}$\quad$ 
      The source code of the original ``main'' program
\item {\tt main}$\quad$
      The parsed document corresponding to the above, such that {\tt b:compile} applied to {\tt main.b} yields {\tt main}.
\end{itemize}

We do this by {\em anticipating}, during the initialization process,
the {\em meaning} of {\tt b:compile} so that we can perform that
action (parsing) on {\tt b:compile.b} to produce {\tt b:compile}. This
can only be achieved by fiat and this is the essence of bootstrapping.

After this minimal initialization, we can then exclusively use the web
interface to develop and extend Wikiplia.

\begin{figure*}[t]
\begin{center}
% \includegraphics[width=0.92\textwidth]{ss_history}
\end{center}
\caption{Screenshot of the history information for {\tt main-go.w}.
} \label{fig:sshistory}
\end{figure*}

@@

As of writing, Wikiplia is at revision~\currentrevision, and has
a number of features implemented.

\subsection{Interface}

The editing interface implemented has been enhanced greatly;
Figure~\ref{fig:sshomepage} shows a screenshot of the main page. The
various views of a key are shown with a series of tabs at the top of
each page. Each user has a home page named after his IP
address,\z\wcite{IP address} which he can use to catalogue his
interests. Various warnings help the user, for example, if he tries to
edit a page that was generated by compiling some source code, a
warning message suggests that he may wish to edit the source code
instead. An ornate logo in SVG\wcite{Scalable Vector Graphics} adorns
the page, and a sidebar provides quick access to the site's features.
The logo and graphics for the site are stored in the repository; the
new {\tt raw} and {\tt typed} actions allows access to these resources
over HTTP so that they may be freely modified.\z\footnote{Because
Wikiplia is dogmatically forward-looking, the graphics require SVG
support in the browser and support for the {\tt data:} URL format, a
combination only found in the newest versions of the Mozilla Firefox.
Wikiplia is compatible with incompatible browsers, however, simply
displaying a crapified version of each page.} At revision 228 support
for metadata was added for each page; a {\tt history} tab now shows
the date, revision number, IP address, and edit summary for each
change to a key (Figure~\ref{fig:sshistory}).

\subsection{Wikiplia language {\tt w}}

Writing XESP documents by hand is very tedious, so one of the first
orders of business was to develop a compiler for a new language, {\tt
w}, which can be extended with convenient features. The first version
of this language, created in revision 22, was written in XESP. It
automatically quoted the appropriate arguments to the {\sf let}, {\sf
it}, {\sf lambda}, {\sf xcase} primitives. In revision 361, the
compiler was ported to the language {\tt w} as {\tt w:compile.w}, then
compiled with the existing {\tt b}-compiled version of {\tt
w:compile.b}, then recompiled with itself until reaching a fixed
point.\z\wcite{Fixed point (mathematics)} After that, more features
were added: a multi-argument function construct {\tt fn}; a
list-deconstructing binding construct {\tt lets}; simple support for
separately-compiled libraries\wcite{Library (computing)} via {\tt
include}; support for mutually-recursive\wcite{Recursion (computer
science)} bundles of functions via {\tt fun};\footnote{Prior to
this, recursion had to be encoded directly by passing an initial
``self'' argument to each function.} and the {\tt cond} keyword for
series of chained ``if\ldots else'' conditionals. At each stage, the
feature is implemented using the current version of {\tt w}, and then
{\tt w:compile.w} is rewritten to use that convenient extension, and
then recompiled until reaching a fixed point.

\subsection{Wikiplia language {\tt page}}

Wikiplia is not just a programming language; it needs facilities for
editing pages that are human readable as well. This can be used to
edit documentation for programming languages, to modify the home page
to tout new developments, to modify the software license
(Section~\ref{sec:license}) or to deface other user's home pages. For
this we provide a Wiki-like\wcite{Wiki} syntax that allows for the
authorship of such pages and easy linking between them. Like other
Wikis, a link to a page that does not exist is colored red, to alert
the user to the opportunity to stake out cyberspace real estate. This
syntax is compiled to XESP documents via the {\tt page} language; the
resulting documents are active in the sense that they check the status
of linked pages to report the correct color on every page load.
